A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-09-10 main.tex
1 tweak

In order to clarify and emphasize the importance of the subtle distinction between population structure and environmental heterogeneity, we present a detailed review of relevant concepts and methods. The remainder of this paper is organized as follows. In section \ref{sec:background} we provide a summary of the causes and consequences of structure in the genetic data of seemingly unrelated individuals and formally define terminology used throughout this paper. We emphasize that confounding due to population structure, frequently cited as a driver of spurious associations, is a function of environmental heterogeneity, where genetic data serves as a proxy for differential environmental exposures \citep{Sillanpaeae2011, sul2018population, vilhjalmsson2012nature, barton2019population}. In so doing, we refine the definition of confounding due to population structure in terms of a non-genetic mechanism. In addition, we briefly review historical methods of correcting for population stratification and relatedness. In section \ref{sec:methods} we provide a detailed review of the two most prevalent multivariate methods of adjusting for population stratification and relatedness at present: PCA adjusted Lasso penalized regression (PC-Lasso) and Lasso penalized multivariate LMMs (LMM-Lasso). We review the statistical details of these methods as well as the consequences \pb{do we review \emph{consequences} in section 3?} \anna{tbd} of the ways in which they attempt to correct for population structure and environmental heterogeneity. In section \ref{sec:results}, we illustrate the concepts described in section \ref{sec:methods} via simulation studies, where the data-generating mechanism formally differentiates between population structure and environmental heterogeneity. In addition, these results identify scenarios where particular methods may be most and least effective in accurately estimating the effect sizes of observed SNPs in the presence of unobserved confounding.

>>>>

In order to clarify and emphasize the importance of the subtle distinction between population structure and environmental heterogeneity, we present a detailed review of relevant concepts and methods. The remainder of this paper is organized as follows. In section \ref{sec:background} we provide a summary of the causes and consequences of structure in the genetic data of seemingly unrelated individuals and formally define terminology used throughout this paper. We emphasize that confounding due to population structure, frequently cited as a driver of spurious associations, is a function of environmental heterogeneity, where genetic data serves as a proxy for differential environmental exposures \citep{Sillanpaeae2011, sul2018population, vilhjalmsson2012nature, barton2019population}. In so doing, we refine the definition of confounding due to population structure in terms of a non-genetic mechanism. In addition, we briefly review historical methods of correcting for population stratification and relatedness. In section \ref{sec:methods} we provide a detailed review of the two most prevalent multivariate methods of adjusting for population stratification and relatedness at present: PCA adjusted Lasso penalized regression (PC-Lasso) and Lasso penalized multivariate LMMs (LMM-Lasso). We review the statistical details of how these methods attempt to correct for population structure and environmental heterogeneity. In section \ref{sec:results}, we illustrate the concepts described in section \ref{sec:methods} via simulation studies, where the data-generating mechanism formally differentiates between population structure and environmental heterogeneity. In addition, these results identify scenarios where particular methods may be most and least effective in accurately estimating the effect sizes of observed SNPs in the presence of unobserved confounding.
